Realizing temporal graphs from fastest paths II

### 2023-10-05 STACS2023/STACS23-shortversion.tex
Some changes, added new meeting whiteboard

\bibliographystyle{plainurl}% the mandatory bibstyle
\usepackage{cite} %order citations by increasing number
\title{Temporal graph realization from fastest temporal paths} %TODO Please add
%%%Realizing temporal graphs from fastest travel times

@@

    Having computed $G$, our algorithm proceeds as follows. We pick an arbitrary edge $e$ and give it label one, that is, $\lambda(e)=1$. Now we push all edges incident with $e$ into a (initially empty) queue. Now we repeat the following as long as the queue is not empty:
\begin{itemize}
    \item Pop edge $e=\{u,v\}$ from the queue. Since $e$ was pushed into the queue, there is an edge $e'$ incident with $e$ that already obtained a label. Let w.l.o.g.\ $e'=\{v,w\}$. Then we set $\lambda(e)=(\lambda(e')-D_{u,w}+1)\bmod \Delta$.
    \item Push all edges incident with $e$ that have not received a label yet into the queue.
\end{itemize}
When the queue is empty, all edges have received a label. Iterate over all vertex pairs $u,v$ and check whether the fastest path from $u$ to $v$ in $(G,\lambda)$ has duration $D_{u,v}$. If this check succeeds for all vertex pairs, output the labelling $\lambda$, otherwise abort.

>>>>

\bibliographystyle{plainurl}% the mandatory bibstyle
\usepackage{cite} %order citations by increasing number
\title{Temporal graph realization from fastest paths} %TODO Please add
%%%Realizing temporal graphs from fastest travel times

@@

    Having computed $G$, our algorithm proceeds as follows. We pick an arbitrary edge $f$ and give it label one, that is, $\lambda(f)=1$. Now we push all edges incident with $f$ into a (initially empty) queue. Now we repeat the following as long as the queue is not empty:
\begin{itemize}
    \item Pop edge $e=\{u,v\}$ from the queue. Since $e$ was pushed into the queue, there is an edge $e'$ incident with $e$ that already obtained a label. Let w.l.o.g.\ $e'=\{v,w\}$. Then we set $\lambda(e)=(\lambda(e')-D_{u,w}+1)\bmod \Delta$.
    \item Push all edges incident with $e$ that have not received a label yet into the queue.
\end{itemize}
When the queue is empty, all edges have received a label. Iterate over all vertex pairs $u,v$ and check whether the fastest path from $u$ to $v$ in $(G,\lambda)$ has duration $D_{u,v}$. If this check succeeds for all vertex pairs, output the labelling $\lambda$, otherwise abort.

### 2023-10-05 STACS2023/abstract.tex
Some changes, added new meeting whiteboard

In this paper we initiate the study of the temporal graph realization problem with respect to the fastest path durations among its vertices, while we focus on periodic temporal graphs. Given an $n \times n$ matrix $D$ and a $\Delta \in \mathbb{N}$, the goal is to construct a $\Delta$-periodic temporal graph with $n$ vertices such that the duration of a fastest path from $v_i$ to $v_j$ is equal to $D_{i,j}$, or to decide that such a temporal graph does not exist. The variations of the problem on static graphs has been well studied and understood since the 1960's, and this area of research remains active until nowadays.

As it turns out, the periodic temporal graph realization problem has a very different computational complexity behavior than its static (i.e. non-temporal) counterpart. First we show that the problem is NP-hard in general, but polynomial-time solvable if the so-called underlying graph is a tree or a cycle. Building upon those results, we investigate its parameterized computational complexity with respect to structural parameters of the underlying static graph which measure the ``tree-likeness''. For those parameters, we essentially give a tight classification between parameters that allow for tractability (in the FPT sense) and parameters that presumably do not. We show that our problem is W[1]-hard when parameterized by the feedback vertex number of the underlying graph, while we show that it is in FPT when parameterized by the feedback edge number of the underlying graph.

>>>>

In this paper we initiate the study of the temporal graph realization problem with respect to the fastest path durations among its vertices, while we focus on periodic temporal graphs. Given an $n \times n$ matrix $D$ and a $\Delta \in \mathbb{N}$, the goal is to construct a $\Delta$-periodic temporal graph with $n$ vertices such that the duration of a fastest path from $v_i$ to $v_j$ is equal to $D_{i,j}$, or to decide that such a temporal graph does not exist. The variations of the problem on static graphs has been well studied and understood since the 1960's (e.g. [Erdos and Gallai, 1960], [Hakimi and Yau, 1965]).

As it turns out, the periodic temporal graph realization problem has a very different computational complexity behavior than its static (i.e. non-temporal) counterpart. First we show that the problem is NP-hard in general, but polynomial-time solvable if the so-called underlying graph is a tree. Building upon those results, we investigate its parameterized computational complexity with respect to structural parameters of the underlying static graph which measure the ``tree-likeness''. We prove a tight classification between such parameters that allow fixed-parameter tractability (FPT) and those which imply W[1]-hardness. We show that our problem is W[1]-hard when parameterized by the feedback vertex number (and therefore also any smaller parameter such as treewidth, degeneracy, and cliquewidth) of the underlying graph, while we show that it is in FPT when parameterized by the feedback edge number (and therefore also any larger parameter such as maximum leaf number) of the underlying graph.

### 2023-10-05 TGR Extended - New paper/Realizing TG - extended.tex
Some changes, added new meeting whiteboard

\subsection{The underlying graph is a path}
We start by investigating the complexity of the problem \DeltaUpperBoundLong\ in the case when the underlying graph of the input instance is a path.
First we observe some properties of the problem.

Let $(G,D)$ be an instance of \deltaUpperBound\, where $G$ is a path of length $n$.
Denote with $v_0,v_1, v_2, \dots, v_n$ the vertices of $G$ where $v_{i}v_{i+1} \in E(G)$, for all $i \in [n]$.
And let $\lambda$ be a labelling that we are trying to construct (\ie a labeling satisfying conditions of the problem).
The following must hold.
\begin{enumerate}[(a)]
	\item \label{path:enum-1}
 Let $i,j,k \in [n]$, where $i < k < j$, and let $D_{i,j} \leq D_{i,k}$. 
 Then the temporal path $P_{i,k}$ from $v_i$ to $v_k$ cannot be of duration more than $D_{i,j} - (k-j)$,
 which implies that $D_{i,k} \leq D_{i,j} - (k-j)$.

@@

From now on we assume that $D$ is a reduced matrix, that satisfies the above conditions, and that the labeling $\lambda$ of $G$, if it exists, does not label two consecutive edges using the same label.

@@

\begin{proof}
We are given an instance  $(G,D)$ of \deltaUpperBound\, where $G$ is a path on vertices $v_0, v_1, \dots, v_{n+1}$. 
Let $\lambda$ be a labeling satisfying this instance.
We denote with $\tau_i^+$ the waiting time at vertex $v_i$ when passing from vertex $v_{i-1}$ to $v_{i+1}$ in $(G,\lambda)$,
and with $\tau_i^-$ the waiting time at $v_i$ when traversing through $v_i$ from the other direction (for details see~\cref{def:travelDelays}).
Since $\tau_i^- = \Delta - \tau_i^+$, we can simplify the above notation and replace $\tau_i^+$ with $\tau_i$, and $\tau_i^-$ with $ \Delta - \tau_i$.
Therefore, by the definition, it follows that the duration of a fastest path among two vertices $v_i, v_j$ in $(G,\lambda)$, for $i < j$
is 
\begin{equation}\label{eq:path-positive}
    d(v_i,v_j)=\sum_{k = i+1}^{j-1} \tau_k + 1,
\end{equation}
and for $i > j$ is 
\begin{equation}\label{eq:path-negative}
d(v_i,v_j)=\sum_{k = i+1}^{j-1} (\Delta - \tau_k) + 1.
\end{equation}

@@

From observing vertices $v_i$ and $v_j$, we get the following two inequalities:
\begin{equation}
\sum_{k = i+1}^{j-1} \tau_k + 1 \leq D_{i,j}
\Leftrightarrow
\sum_{k = i+1}^{j-1} \tau_k \leq D_{i,j} - 1
\end{equation}
and
\begin{equation}
\sum_{k = i+1}^{j-1} (\Delta - \tau_k) + 1
\Leftrightarrow
\sum_{k = i+1}^{j-1} - \tau_k \leq  D_{i,j} - 1 -(j-i+1)\Delta .
\end{equation}
From the condition $\tau_i \in [\Delta]$, we add the inequality $\tau_i \leq \Delta$ and $\tau_i \geq 0$.

Let us now describe our LP.
We want to find an assignment of values for each $\tau_i$, that satisfy the inequalities:
\begin{equation} \label{eq:paths-inequalities-A}
\begin{aligned}
\tau_1 & \leq \Delta \\
\tau_2 & \leq \Delta \\
& \vdots \\
\tau_n &\leq \Delta \\
%%%% \tau \leq Delta
\tau_1 &\leq D_{0,2} -1 \\
\tau_1 + \tau_2 &\leq D_{0,3} -1 \\
\tau_1 + \tau_2 + \tau_3 &\leq D_{0,4} -1 \\
& \vdots \\
\tau_1 + \cdots + \tau_n &\leq D_{0,n+1} -1 \\
%%%% all with \tau_1
\tau_2 & \leq D_{1,3} -1 \\
\tau_2 + \tau_3 &\leq D_{1,4} - 1\\
& \vdots \\
\tau_n &\leq D_{n-1,n+1} - 1 \\
%%% all positive tau
%%%
- \tau_1 &\leq D_{2,0} -1 - \Delta \\
- \tau_1 - \tau_2 &\leq D_{3,0} -1 -2 \Delta \\
& \vdots \\
- \tau_n &\leq D_{n+1,n-1} - 1 - \Delta
\end{aligned}
\end{equation}
where all $\tau_i$ are positive integers.
We can write the above set of inequalities in the matrix form as $A t \leq b$, where $t$ is a vector of variables $\tau_1,\tau_2,\dots, \tau_n$.
Therefore we can form the following LP, in the canonical form
\begin{equation} \label{eq:paths-LP}
\begin{aligned}
\text{find a vector} &&& t \\
\text{that maximizes} &&& \sum_i \tau_i \\
\text{subject to} && A t  &\leq b, \\
\text{and} && t &\geq 0. \\
\end{aligned}
\end{equation}
Note that in our case we do not necessarily want to maximize the sum of all $\tau_i$ variables, but just to find an assignment of variables that will satisfy our conditions. Clearly if there is a solution to our system then there is also a solution that will maximize the sum, and vice versa.

>>>>

\subsection{The underlying graph is a path}
We start by investigating the complexity of the problem \DeltaUpperBoundLong\ in the case when the underlying graph of the input instance is a path.

\begin{comment}
Let $(G,D)$ be an instance of \deltaUpperBound\, where $G$ is a path of length $n$.
Denote with $v_0,v_1, v_2, \dots, v_n$ the vertices of $G$ where $v_{i}v_{i+1} \in E(G)$, for all $i \in [n]$.
And let $\lambda$ be a labelling that we are trying to construct (\ie a labeling satisfying conditions of the problem).
The following must hold.
\begin{enumerate}[(a)]
	\item \label{path:enum-1}
 Let $i,j,k \in [n]$, where $i < k < j$, and let $D_{i,j} \leq D_{i,k}$. 
 Then the temporal path $P_{i,k}$ from $v_i$ to $v_k$ cannot be of duration more than $D_{i,j} - (k-j)$,
 which implies that $D_{i,k} \leq D_{i,j} - (k-j)$.

@@

From now on we assume that $D$ is a reduced matrix, that satisfies the above conditions, and that the labeling $\lambda$ of $G$, if it exists, does not label two consecutive edges using the same label.
\end{comment}

@@

Before stating the proof let us observe the following.
Suppose that $(G, D)$ is a yes instance of \deltaUpperBound\, where $G$ is a path on vertices $v_0, v_1, v_2, \dots, v_n$, 
and let $\lambda$ be a labeling of $G$ s.t. $(G,\lambda)$ realizes $D$.
Then, w.l.o.g. we can assume that $\lambda(v_{i-1} v_{i}) \neq \lambda (v_{i}v_{i+1})$ for all $i \in [n-1]$, \ie any two consecutive edges are labeled with a different label.
This follows from the fact that the duration among two vertices $v_i, v_{i+2}$ on distance $2$ in $G$ equals to $\Delta + 1$ when the two consecutive labels are the same, 
and the duration is at most $\Delta$,
when the two consecutive labels are not the same,
Therefore, if the value $D_{i,i+2}= \Delta + 1$ and $\lambda(v_{i-1} v_{i}) \neq \lambda (v_{i}v_{i+1})$, we get that the duration $d(v_i, v_{i+2})$ from vertex $v_i$ to $v_{i+2}$ in $(G, \lambda)$ is at most $\Delta$ 
and therefore $d(v_i, v_{i+2}) \leq D_{i,i+2}$,
which still satisfies the conditions of \deltaUpperBound.

From now on we assume that the labeling $\lambda$ of $G$, if it exists, does not label two consecutive edges using the same label.

\begin{proof}
We are given an instance  $(G,D)$ of \deltaUpperBound\, where $G$ is a path on vertices $v_0, v_1, \dots, v_{n+1}$. 
Let $\lambda$ be a labeling satisfying this instance.
We denote with $\tau_i^+$ the waiting time at vertex $v_i$ when passing from vertex $v_{i-1}$ to $v_{i+1}$ in $(G,\lambda)$,
and with $\tau_i^-$ the waiting time at $v_i$ when traversing through $v_i$ from the other direction (for details see~\cref{def:travelDelays}).
Since $\tau_i^- = \Delta - \tau_i^+$, 
\todo[inline]{NK: Done - paragraph above the proof \\
GM: not correct, they can be both Delta ! THIS MUST BE WRITTEN. THE TRICK IS THAT, FOR A PATH, WE WILL NEVER CHOOSE TO HAVE TWO CONSECUTIVE LABELS EQUAL, SINCE THIS WOULD JUST INCREASE THE FASTEST TIMES ARTIFICIALLY. BUT THIS IS NOT TRUE FOR A GRAPH THAT IS NOT A PATH !}
we can simplify the above notation and replace $\tau_i^+$ with $\tau_i$, and $\tau_i^-$ with $ \Delta - \tau_i$.
Therefore, by the definition, it follows that the duration of a fastest path among two vertices $v_i, v_j$ in $(G,\lambda)$, for $i < j$
is 
\begin{equation}\label{eq:path-positive}
    d(v_i,v_j)=\sum_{k = i+1}^{j-1} \tau_k + 1,
\end{equation}
and for $i > j$ is 
\begin{equation}\label{eq:path-negative}
d(v_i,v_j)=\sum_{k = j-1}^{i+1} (\Delta - \tau_k) + 1.
\end{equation}
\todo[inline]{NK: Done \\
GM: WRONG INDICES: CHANGE IT HERE AND BELOW}

@@

From observing vertices $v_i$ and $v_j$, we get the following two inequalities:
\begin{equation}
\sum_{k = i+1}^{j-1} \tau_k + 1 \leq D_{i,j}
\Leftrightarrow
\sum_{k = i+1}^{j-1} \tau_k \leq D_{i,j} - 1
\end{equation}
and
\begin{equation}
\sum_{k = j-1}^{i+1} (\Delta - \tau_k) + 1 \leq D_{j,i}
\Leftrightarrow
\sum_{k = i+1}^{j-1} - \tau_k \leq  D_{j,i} - 1 -(j-i+1)\Delta .
\end{equation}
\todo[inline]{NK: Done \\
GM: MISSING INEQUALITY ON THE LEFT SIDE}
From the condition $\tau_i \in [\Delta]$, we add the inequality $\tau_i \leq \Delta$ and $\tau_i \geq 1$.

Let us now describe our LP.
We want to find an assignment of values for each $\tau_i$, that satisfy the inequalities:
\begin{equation} \label{eq:paths-inequalities-A}
\begin{aligned}
\tau_1 & \leq \Delta \\
\tau_2 & \leq \Delta \\
& \vdots \\
\tau_n &\leq \Delta \\
%%%% \tau \leq Delta
\tau_1 &\leq D_{0,2} -1 \\
\tau_1 + \tau_2 &\leq D_{0,3} -1 \\
\tau_1 + \tau_2 + \tau_3 &\leq D_{0,4} -1 \\
& \vdots \\
\tau_1 + \cdots + \tau_n &\leq D_{0,n+1} -1 \\
%%%% all with \tau_1
\tau_2 & \leq D_{1,3} -1 \\
\tau_2 + \tau_3 &\leq D_{1,4} - 1\\
& \vdots \\
\tau_n &\leq D_{n-1,n+1} - 1 \\
%%% all positive tau
%%%
- \tau_1 &\leq D_{2,0} -1 - \Delta \\
- \tau_1 - \tau_2 &\leq D_{3,0} -1 -2 \Delta \\
& \vdots \\
- \tau_n &\leq D_{n+1,n-1} - 1 - \Delta
\end{aligned}
\end{equation}
where all $\tau_i$ are positive integers.
We can write the above set of inequalities in the matrix form as $A t \leq b$, where $t$ is a vector of variables $\tau_1,\tau_2,\dots, \tau_n$.
Therefore we can form the following LP, in the canonical form
\begin{equation} \label{eq:paths-LP}
\begin{aligned}
\text{find a vector} &&& t \\
\text{that maximizes} &&& \sum_i \tau_i \\
\text{subject to} && A t  &\leq b, \\
\text{and} && t &\geq 0. \\
\end{aligned}
\end{equation}
\todo[inline]{GM: WE DO NOT WANT AN OBJECTIVE FUNCTION, BUT EVEN IF WE WANT, WHY MAX AND NOT MIN? WE WANT HERE FEASIBILITY LP}
Note that in our case we do not necessarily want to maximize the sum of all $\tau_i$ variables, but just to find an assignment of variables that will satisfy our conditions. Clearly if there is a solution to our system then there is also a solution that will maximize the sum, and vice versa.
